Automatic feature classification in stochastic block models

### 2021-04-18 nips/submission.tex
Work on feat-gen

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{bbm}

@@

We are presented with a vertex-labelled graph $(A, X)$. The goal is to draw samples from equation \ref{eqn:theta-target}. However, this is not easily done in practice.
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(i)} \sim p(\theta | A, X)
\end{equation}
%
We instead propose an iterative approach. First drawing samples $b^{(i)}$ from the block membership posterior (equation \ref{eqn:b-samples}). We then use each $b^{(i)}$ to draw samples for $\theta$ as in equation \ref{eqn:theta-samples}. 
%
\begin{align}
	b^{(i)} &\sim p \Big( b | A, X \Big)  \label{eqn:b-samples}\\
	\theta^{(i)} &\sim p\Big(\theta | X, b^{(i)} \Big) \label{eqn:theta-samples}
\end{align}
%
Both of these can be implemented with a Markov Chain through the Metropolis-Hastings algorithm \cite{hastings-alg}. We just need to define a proposal distribution $q(x, y)$ for proposing a move $x \rightarrow y$ and be able to evaluate an un-normalised form of the target distribution, denoted $\pi(\cdot)$, point-wise. The proposed move is then accepted with acceptance probability $\alpha$ else it is rejected and we stay at $x$.
%
\begin{equation}
	\alpha = \min \left( \frac{\pi(y) q(y, x)}{\pi(x) q(x, y)} , 1 \right)
\end{equation}
%
This accept-reject step ensures the resulting Markov Chain is in detailed balance with the target distribution $\pi(\cdot)$. What we propose in equations \ref{eqn:b-samples} and \ref{eqn:theta-samples} is therefore implemented through a 2-level Markov chain. The resulting samples for $\theta^{(i)}$ are unbiased in the sense that the expectation of their distribution is the posterior we are targeting in equation \ref{eqn:theta-target}.
%
\begin{align*}
	\Expect_{b^{(i)}} \left[p \left( \theta | X, b^{(i)} \right) \right] &= \sum_{b \in \mathcal{B}^N} p(\theta | X, b) p(b | A, X) \\
	&= \sum_{b \in \mathcal{B}^N} p(\theta, b | A, X) \\
	&= p(\theta | A, X) \\
\end{align*}
%
Which is indeed the distribution we are targeting from equation \ref{eqn:theta-target}. The reason we split the Markov chain into two stages is because the summation over all latent states $b \in \mathcal{B}^N$ required to compute the likelihood $p(A| X, \theta) = \sum_{b \in \Bcal^N}$ is computationally intensive

\subsection{Inferring block memberships}

\citet{Peixoto-MCMC} proposes a Monte Carlo method which we will base our approach on. It relies on writing the posterior in the following form.
%
\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) = \pi_b(b)
\end{equation}
%
Now $\pi_b(\cdot)$ is the un-normalised density we wish to sample from. We can break $\pi_b$ down as follows:
%
\begin{align*}
	\pi_b(b) &= p(b|X) \sum_{\psi} \nolimits p(A , \psi | b, X) \\
	&= p(b|X) p(A, \psi^* | b, X) \\
	&= p(A | b, \psi^*) \cdot p(\psi^* | b) \cdot p(b | X)
\end{align*}
%
Since we are using a microcanonical formulation, there is only one value of $\psi$ which we denote $\psi^*$ that is compatible with the given $A$ and $b$. Namely, $k^*_i = \sum_j A_{ij} $ and $e^*_{rs} = \sum_{i, j} A_ij \one \{b_i=r\} \one\{b_j=s\}$. Therefore, the summation over all $\psi$ reduces to just one term that has the value $\psi^*$ compatible with $b$ and $\theta$. We also define the description length or entropy of the configuration as.
%
\begin{equation}
	S(b) = - \log \pi_b(b) = - \left[ \log p(A | b, \psi^*) + \log p(\psi^*, b | X) \right]
\end{equation}
%
Now this entropy can be thought of as the description length of the graph $A$ because it is the sum of the information required to represent the graph given the parameters and the amount of information required to store the parameters given $X$.

@@

@@

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{bbm}
\usepackage{amssymb}

@@

We are presented with a vertex-labelled graph $(A, X)$. The goal is to draw samples from equation \ref{eqn:theta-target}. However, this is not easily done in practice.
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(i)} \sim p(\theta | A, X)
\end{equation}
%
We instead propose an iterative approach. First drawing samples $b^{(i)}$ from the block membership posterior (equation \ref{eqn:b-samples}). We then use each $b^{(i)}$ to draw samples for $\theta$ as in equation \ref{eqn:theta-samples}. 
%
\begin{align}
	b^{(i)} &\sim p \Big( b | A, X \Big)  \label{eqn:b-samples}\\
	\theta^{(i)} &\sim p\Big(\theta | X, b^{(i)} \Big) \label{eqn:theta-samples}
\end{align}
%
Both of these can be implemented with a Markov Chain through the Metropolis-Hastings algorithm \cite{hastings-alg}. We just need to define a proposal distribution $q(x, y)$ for proposing a move $x \rightarrow y$ and be able to evaluate an un-normalised form of the target distribution, denoted $\pi(\cdot)$, point-wise. The proposed move is then accepted with acceptance probability $\alpha$ else it is rejected and we stay at $x$.
%
\begin{equation}
	\alpha = \min \left( \frac{\pi(y) q(y, x)}{\pi(x) q(x, y)} , 1 \right)
\end{equation}
%
This accept-reject step ensures the resulting Markov Chain is in detailed balance with the target distribution $\pi(\cdot)$. What we propose in equations \ref{eqn:b-samples} and \ref{eqn:theta-samples} is therefore implemented through a 2-level Markov chain. The resulting samples for $\theta^{(i)}$ are unbiased in the sense that the expectation of their distribution is the posterior we are targeting in equation \ref{eqn:theta-target}.
%
\begin{align*}
	\Expect_{b^{(i)}} \left[p \left( \theta | X, b^{(i)} \right) \right] &= \sum_{b \in \mathcal{B}^N} p(\theta | X, b) p(b | A, X) \\
	&= \sum_{b \in \mathcal{B}^N} p(\theta, b | A, X) \\
	&= p(\theta | A, X) \\
\end{align*}
%
Which is indeed the distribution we are targeting from equation \ref{eqn:theta-target}. The reason we split the Markov chain into two stages is because the summation over all latent states $b \in \mathcal{B}^N$ required to directly compute the likelihood $p(A| X, \theta) = \sum_{b \in \Bcal^N} p(A | b) P(b | X, \theta)$ is computationally intensive.

\subsection{Sampling block memberships}

\citet{Peixoto-MCMC} proposes a Monte Carlo method which we will base our approach on. It relies on writing the posterior in the following form.
%
\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) = \pi_b(b)
\end{equation}
%
Now $\pi_b(\cdot)$ is the un-normalised density we wish to sample from. In other words, we wish to construct a Markov chain that has $\pi_b(\cdot)$ as its invariant distribution. We can break $\pi_b$ down as follows:
%
\begin{align*}
	\pi_b(b) &= p(b|X) \sum_{\psi} \nolimits p(A , \psi | b, X) \\
	&= p(b|X) p(A, \psi^* | b, X) \\
	&= p(A | b, \psi^*) \cdot p(\psi^* | b) \cdot p(b | X)
\end{align*}
%
Since we are using a microcanonical formulation, there is only one value of $\psi$ which we denote $\psi^*$ that is compatible with the given $(A, b)$ pair. Specifically, $k^*_i = \sum_j A_{ij} $ and $e^*_{rs} = \sum_{i, j} A_ij \one \{b_i=r\} \one\{b_j=s\}$. Therefore, the summation over all $\psi$ reduces to just the single $\psi^*$ term. We also define the microcanonical entropy of the configuration as.
%
\begin{equation}
	S(b) = - \log \pi_b(b) = - \left[ \log p(A | b, \psi^*) + \log p(\psi^*, b | X) \right]
\end{equation}
%
This entropy can be thought of as the description length of the graph because it is the sum of the information required to represent the graph given the parameters and the amount of information required to store the parameters (given the feature matrix $X$).

@@

Now the invariant distribution we wish to target for the $\theta$ samples is the posterior of $\theta$ given the values of the pair $(X, b)$. We write this as follows:
%
\begin{align}
	p(\theta | X, b) &\propto p(b | X, \theta) p(\theta) = \pi_\theta (\theta) =  \exp \left( - U(\theta) \right) \\
	\therefore U(\theta) &= - \left( \log p(b | X, \theta) + \log (\theta) \right)
\end{align}
%
Where we have introduced $U(\theta)$ equal to the negative log posterior, because it simplifies analysis. Each of these terms is easily computed, by first defining $t_{ij} \coloneqq \one \left\{ b_i = j \right\}$ and $y_{ij} = \phi_j(x_i; \theta)$.
%
\begin{equation}
	\log p(b | X, \theta) = \sum_{i=1}^{N} \sum_{j=1}^{B} t_{ij} \log y_{ij}  \quad \textrm{and} \quad
	\log p(\theta) = -\frac{d}{2} \log 2\pi - \frac{1}{2} || \theta || ^2
\end{equation}
%
The proposal can then be written as:
%
\begin{equation}
	\theta' = \theta - h_t \nabla U(\theta) + \sqrt{2h_t} \cdot \xi_t
\end{equation}
%
Where $\xi_t \sim \Gaussian(0, I)$. As such the proposal distribution, is easy to compute
%
\begin{equation}
	q(\theta, \theta') = \Gaussian \left( \theta' ; \theta - h_t \nabla U(\theta), 2h_t I \right)
\end{equation}
%
The term $\nabla U$ has an easy to compute analytic form. By noting that $\theta = \{w_k\}_{k=1}^{B}$
%
\begin{equation}
	\frac{\partial U}{\partial w_k} = - \left( \sum_{i=1}^{N} \Big\{ \tilde{x}_i (t_{ik} - y_{ik}) \Big\} - w_k \right)
\end{equation}
%
. We would then accept these samples with probability
%
\begin{equation}
	\alpha = \min \left( \exp \left(-\Delta U \right) \frac{ q(\theta', \theta)}{q(\theta, \theta')} , 1 \right)
\end{equation}

@@

\subsection{Derivation of gradient with respect to feature parameters}
